### 2019-01-11 writing/thesis/chapters/methods.tex
Update methods.

% Stuff for numerics
\newcommand{\domain}{\Omega}
\newcommand{\broken}{\domain}
\newcommand{\cell}[1][]{C_{#1}}
\newcommand{\boundary}{\partial \domain}
\newcommand{\lagrange}[1][i]{\varphi_{#1}}
\newcommand{\lagrangeRef}[1][i]{\hat{\varphi}_{#1}}
\NewDocumentCommand{\sbasis}{ O{} O{} }{\phi^{#1}_{#2}}
\NewDocumentCommand{\stbasis}{ O{} O{} }{ \Theta^{#1}_{#2}}
\NewDocumentCommand{\sbasisRef}{ O{} O{} }{\hat{\phi}^{#1}_{#2}} %{\hat{\sbasis[#1][#2]}}
\NewDocumentCommand{\stbasisRef}{ O{} O{} }{\hat{\Theta}^{#1}_{#2}} %{\hat{\stbasis[#1][#2]}}
%\newcommand{\stestfunction}[1]{\sbasis[#1]}
\NewDocumentCommand{\stestfunction}{ O{} O{} }{\sbasis[#1][#2]}
\NewDocumentCommand{\sttestfunction}{ O{} O{} }{\stbasis[#1][#2]}
\newcommand{\normal}{\bm{n}}
\newcommand{\dsol}[1][h]{\bm{u}_{#1}}

@@

@@

% Integrals
\newcommand{\curTime}{t^n}
\newcommand{\nextTime}{t^{n+1}}
\newcommand{\intdt}[1]{\int_{\curTime}^{\nextTime} #1 \dd{t}}
\newcommand{\intdcell}[1]{\int_{\cell} #1 \dd{\bm{x}}}
\newcommand{\intdrefcell}[1]{\int_{\cell} #1 \dd{\hat{\bm{x}}}}
\newcommand{\intdcellb}[1]{\int_{\partial{} \cell} #1 \dd{S}} % TODO: define boundary of cell
\newcommand{\intdrefcellb}[1]{\int_{\partial{} \cell} #1 \hat{\dd{S}}} % TODO: define boundary of cell

% Yet to be sorted
\newcommand{\quadWeight}[1][i]{w_{#1}}
\newcommand{\quadNode}[1][i]{n_{#1}}
\newcommand{\sbasisSize}{(N+1)^\dimensions}
\newcommand{\stbasisSize}{(N+1)^{\dimensions+1}}

\chapter{An ADER-DG scheme for the Navier-Stokes Equations}\label{chap:methods}
This chapter describes the numerical and physical background that is needed to simulate reacting fluids with good accuracy.
We start with a description 
We start with a description of the arbitrary high order derivatives (\ader) discontinuous Galerkin (\dg) method with a focus on equations that contain both advective and diffusive terms.
We then seque into a short description of the \muscl\ scheme, a second order finite volume scheme.

We conclude this chapter by a description of adaptive mesh refinement (\amr) and finite volume limiting, which combines both numerical methods to achieve a stable, high-order method.

\section{Reactive Compressible Navier Stokes Equations}\label{sec:navier-stokes}
\newcommand{\diffCoeff}{\varepsilon}
\newcommand{\hyperFluxDef}{
  \begin{pmatrix}
    \Qj \\
    \Qv  \otimes \Qj + \bm{I} \pressure  \\
    \Qv \cdot (\bm{I} \QE + \bm{I} \pressure) \\
    \Qj \QZZ
  \end{pmatrix}
}

\newcommand{\viscFluxDef}{
  \begin{pmatrix}
     -\diffCoeff \gradient{\Qrho}\\
     \stressT (\Q, \gradQ)  \\
     \Qv \cdot \stressT (\Q, \gradQ) - \kappa \gradient{T}\\
     -\diffCoeff \gradient{\QZ}
   \end{pmatrix}
}

Fluid motion can be described by the compressible Navier Stokes equations.
We follow the description in~\cite{dumbser2010arbitrary} for the Navier Stokes part.
The coupling with the advection-diffusion-reaction equation follows~\cite{hidalgo2011ader} which describes this equation set for the one-dimensional case.
In the following, we will denote the spatial coordinates as $\bm{x} = \left( x,z \right)$ and $\bm{x} = \left( x,y,z \right)$ for the two and three-dimensional case respectively.
The variable $t$ corresponds to the currenty (physical) time.
We solve the \pde\ in the conversative form \pcref{eq:conservation-law-gradient}\todo{Dont reference to following chapter!}
\begin{equation}
 \label{eq:equation-set} 
 \begin{array}{l}
 \text{mass cons.} \\
 \text{momentum cons.} \\
 \text{energy cons.} \\
 \text{cont.\ gas} 
\end{array}
:
\quad
  \pdv{}{t}
  \underbrace{
  \begin{pmatrix}
    \Qrho\\
    \Qj\\
    \QE\\
    \QZ
    \end{pmatrix}}_{\Q}
  +
  \divergence{
  \underbrace{
  \left(
   \underbrace{\hyperFluxDef}_{\hyperFlux(\Q)}
+
\underbrace{\viscFluxDef}_{\viscFlux(\Q, \gradQ)}
  \right)}_{\flux(\Q, \gradQ)}}
 =
  \underbrace{
  \begin{pmatrix}
    \source[\Qrho\phantom{\Qrho}]\\
    \source[\Qj]\\
    \source[\QE]\\
    \source[\QZ]
    \end{pmatrix}}_{\source(\Q, \bm{x}, t)}
\end{equation}
which describes the time evolution of variables $\Q$ with respect to a flux $\flux(\Q, \gradQ)$ and a source $\source(\Q)$.
The vector of conserved quantities is given by
\begin{equation}
  \label{eq:conserved-variables}
 \Q = \left( \Qrho, \Qj, \QE, \QZ \right),
\end{equation}
where $\Qrho$ is the density, $\Qj$ is the two or three-dimensional momentum, $\QE$ the energy density and $\QZ$ is the mass fraction of the chemical reactant.
The chemical reactant in our case is unburnt gas.

The rows of the equation are the conservation of mass, the conservation of momentum, the conservation of energy and the continuity of the reactant.
We split the flux into a hyperbolic $\hyperFlux(\Q)$ and a viscous part$\viscFlux(\gradQ)$

\begin{equation}
  \label{eq:flux}
  \flux(\Q, \gradQ) = \hyperFlux(\Q) + \viscFlux(\Q, \gradQ).
\end{equation}
The hyperbolic flux is given by
\begin{equation}
  \label{eq:hyper-flux}
  \hyperFlux(\Q) = \hyperFluxDef,
\end{equation}
which are the Euler equations coupled with an advection equation.
In this equation $\bm{a} \otimes \bm{b} = \bm{a} \bm{b}^\intercal$ is the Kronecker or outer product of two vectors $\bm{a}$ and $\bm{b}$.
The pressure $\pressure$ is given by the caloric equation of state of an ideal reacting gas\todo{Geopot height correct?}
\begin{equation}
  \label{eq:eos}
  \pressure = (\gamma - 1) \left(\QE - \frac{1}{2} \left(\Qv \cdot \Qj \right)  - q_0 \QZ - gz \right).
\end{equation}
The term $q_0 \QZ$ is the chemical energy with the heat release $q_0$, the term $gz$ is the geopotential height.
The pressure is related to the temperature $T$ by the thermic ideal gas law
\begin{equation}
  \label{eq:temperature}
 \frac{\pressure}{\Qrho} = RT,
\end{equation}
where $R$ is the specific gas constant of a fluid.

@@

We also support two different kind of (optional) source terms.
\todo{Either include gravity in pressure or add source term for energy.}
The first source term is gravity, given by
\begin{equation}\label{eq:source-gravity}
  \source[\Qj] = - g \bm{k} \cdot \Qj,
\end{equation}
where $g = 9.81$ and $\bm{k}$ is a unit vector pointing in $z$-direction.

\newcommand{\backgroundPressure}{\overline{\pressure}}
\newcommand{\backgroundRho}{\overline{\Qrho}}
Some of our scenarios can be described as a pertubation of a background state that is in hydrostatic balance
\begin{equation}
  \label{eq:hydrostatic-balance}
  \pdv{}{z} \backgroundPressure{\left (z \right )} = -g \backgroundRho(z),
\end{equation}
where $\backgroundPressure$ and $\backgroundRho$ is the background pressure and density.
We now focus on the following part of the momentum equation (neglecting viscous effects) of \cref{eq:equation-set}
\begin{equation}
  \label{eq:momentum-equation}
  \pdv{\Qj}{t}+ \divergence{
    \Qv \cdot \left( \bm{I} \QE + \bm{I} \pressure \right)
  }
  =
  -g \bm{k} \cdot \Qrho
\end{equation}.
The background state \cref{eq:hydrostatic-balance} leads to large values and can diminish other parts of the flux due to numerical instabilities~\cite{muller2010adaptive}.
We can now split the pressure as $\pressure = \backgroundPressure(z) + \pressure' (\bm{x}, t)$ and the density as $\Qrho = \backgroundRho(z) + \Qrho'(\bm{x}, t)$.
Inserting this splitting, the definition of divergence into \cref{eq:momentum-equation} allows us to apply \cref{eq:hydrostatic-balance}.
Furthermore, the background states have zero derivatives in all directions but in $z$.
We arrive at 
\begin{equation}
  \label{eq:momentum-equation-split}
  \pdv{\Qj}{t}+ \divergence \left(
    \Qv \cdot \left( \bm{I} \QE + \bm{I} \pressure' \right)
   \right)
  =
  -g \bm{k} \Qrho'.
\end{equation}.
We thus cancel parts of the flux with parts of the source term.
Note that the time derivative of the background density also vanishes.
We do not use this to simplify the implementation of the gradient computation.
This form of the equation is similar to equation set 3 of~\cite{giraldo2008study} but without splitting the energy.

\newcommand{\reactionTimescale}{\tau}
\newcommand{\reactionTemperature}{T_{\text{ign}}}
The chemical reactive source term is
\begin{equation}\label{eq:source-reaction}
  \source[\QZ] = - K(T) \QZ.
\end{equation}
We use the very simple reaction model
\begin{equation}\label{eq:reaction-model}
\begin{cases}
  - \frac{1}{\reactionTimescale} & T > \reactionTemperature\\
  0 & \text{otherwise}
\end{cases}  
\end{equation}
where $\reactionTimescale$ is the timescale of the reaction and $\reactionTemperature$ is the activation temperature.

\subsection{Boundary conditions}
To close the system we need to impose boundary conditions.

@@

We manually set the energy component of the numerical flux to zero for both no- and free-slip conditions.
\section{The ADER-DG Method}\label{sec:ader-dg}
\newcommand{\NVar}{N^{\text{var}}}
\newcommand{\dimensions}{d}
\newcommand{\mapping}{\mathcal{M}}
\newcommand{\volume}{V}
\newcommand{\cellCenter}{\operatorname{cell-center}}
We describe the arbitrary derivative discontinous Galerkin (\aderdg) method in this chapter.
Our description of the main method follows~\cite{dumbser2008unified,dumbser2010arbitrary,dumbser2018efficient}, our notation follows primarily~\cite{dumbser2018efficient}.
Let $\NVar$ be the number of variables and $\dimensions$ the number of spatial dimensions.
We discuss the approximation of systems of partial differential equations (\pde) of the form
\begin{equation}
  \label{eq:conservation-law-gradient}
 \frac{\partial}{\partial_t}  \Q + \div \flux(\Q, \gradQ) = \source(\bm{x}, t, \Q),
\end{equation}
which in contrast to hyperbolic conservation laws of the form
\begin{equation}
  \label{eq:conservation-law}
 \frac{\partial}{\partial_t}  \Q + \div \flux(\Q) = \source(\bm{x}, t, \Q)
\end{equation}
contain the gradient $\gradQ \in \mathbb{R}^{\NVar \times \dimensions}$ of the so called vector of conservative variables $\Q \in \mathbb{R}^{\NVar}$~\cite{dumbser2010arbitrary}.
They are therefore not hyperbolic but rather parabolic or elliptic.

We discuss the solution of a conservation law \cref{eq:conservation-law-gradient} with two or three dimensional domain $\domain$ and boundary $\boundary$.
In our implementation of the discontinous Galerkin (\dg) framework, we approximate this solution in the space
\begin{equation}
  \label{eq:dg-space}
  \broken = \bigcup_k \cell[k]
\end{equation}
of disjoint quadrilateral cells $\cell$.
Note that we do not distinguish between the broken approximation space and the domain, the use should be clear given its surrounding context.
For each cell we denote its center by $\cellCenter{\cell}$ and its size by $\Delta x, \Delta y$ and $\Delta z$.
%In the following we make use of the Einstein summation convention where summation over repeated indexes is implied.
% Inside each cell $\cell$ we represent the solution in terms of the basis function 
% \begin{equation}
%   \label{eq:cell-approx}
%   u(\bm{x}, t^n)_{|\cell[i]} = \hat{u}^n_{i,l} \sbasis{l}(\bm{x}),
% \end{equation}
% where $l$ is a multi-index, containing one index per spatial dimension.
% \todo{dofs $\hat{u}$, local solution, etc.}
% For example, $(l = (l_1, l_2))$ for the two dimensional case.
% This polynomial is interpolating, i.e.\ \ldots\todo{interpolating?}.
% This choice of basis (called \textit{nodal} basis) allows us to easily compute integrals over cells using Gaussian quadrature.
% In detail, we use a the Lagrange interpolation polynomials.

@@

\sidetitle{Reference Cell}
For reasons of computational efficiency, we describe the scheme in terms of a reference cell.
We use the quad cell $\hat{\cell} = [0, 1]^\dimensions$ as space reference cell.\todo{check if $[0,1]$ detontes correct interval}.
The mapping, which takes a point on a grid cell with coordinates $\bm{x}$ and returns the corresponding point $\hat{\bm{x}}$ on the reference cell
\begin{equation}\label{eq:space-mapping}
  \hat{x} = \mapping (\bm{x}) =
  \cellCenter(\cell) +
\begin{pmatrix}
\Delta x && \\
&\Delta y & \\
&&\Delta z&
\end{pmatrix}
  \begin{pmatrix}
    x - 0.5\\
    y - 0.5\\
    z - 0.5
  \end{pmatrix}.
\end{equation}
We also define the inverse mapping $\mapping^{-1}$.
This can be used to define a function $f(\bm{x})$ acting on a point $\bm{x}$ of a regular cell in terms of a function $\hat{f}(\bm{x})$ that acts on the corresponding point of the reference cell by the relation
\begin{equation}
  \label{eq:function-ref-cell}
  f(\bm{x}) = f\left( \mapping^{-1}(\bm{x}) \right) = \hat{f}(\hat{\bm{x}}).
\end{equation}

Similar to this, we define a mapping that maps a point in time to the reference time $\hat{t} \in (0,1)$
\newcommand{\timeMapping}{\mathcal{T}}
\begin{equation}
  \label{eq:time-mapping}
  \hat{t} = \timeMapping(t) = t_0 + (\Delta t) t,
\end{equation}
where $\Delta t$ is timestep of a cell and $t_0$ is the beginning of its time interval.
The space-time reference cell is then $\hat{\cell} = [0,1]^{D+1}$

The Jacobian determinant of $\mapping$ is simply the volume $V$ of the cell
\begin{equation}
  \label{eq:determinant-mapping}
  \volume = \operatorname{det}(\operatorname{Jacobian}\mapping) = \Delta x \, \Delta y \,\Delta z,
\end{equation}
the Jacobian determinant of $\timeMapping$ is $\Delta t$.
This is useful for evaluating derivatives by the chain rule and, similarly, integrals by substitution.
For example, we can evaluate a volume integral by
\begin{equation}
  \label{eq:integration-by-substitution}
  \intdcell{
f(\bm{x})
  }
  =
\volume \intdrefcell{
    \hat{f}(\hat{\bm{x}})
  }
\end{equation}
\sidetitle{Basis Functions}
The next step is the definition of basis functions for the reference cell.
We use Langrange polynomials that are defined in the one-dimensional case for the reference interval $(0,1)$ by
\begin{align}
  \label{eq:lagrange-basis}
    \lagrangeRef(x) = \prod_{i \neq n}^N \frac{x - {\quadNode{j}}}{x_i - \quadNode[j]},
\end{align}
We can then approximate functions as a sum over the support points $\quadNode[j]$
\begin{equation}
  \label{eq:langrange-expansion-1d}
  f(x) = \sum_{i = 0}^N f(x_i) \lagrangeRef[i](x) = f(x_i) \lagrangeRef[i](x),
\end{equation}
where we used the Einstein summation convention where summation over repeated indices is implied\todo{Remove this sentence, or sentence before introducing einstein summation!}.
The support points $\quadNode[n]$ are chosen such that they coincide with the nodes of the Gauss-Legendre quadrature of order $N+1$ which we can use to evaluate integrals
\begin{equation}
  \label{eq:quadrature-1d}
  \int_0^1 f(x) \dd{x} = \sum_n^{N+1} \quadWeight[n] f(x_n),
\end{equation}
where $\quadNode[n]$ is a quadrature point with corresponding weight $\quadWeight[n]$.

>>>>

% Domain
\newcommand{\domain}{\Omega}
\newcommand{\boundary}{\partial \domain}
\newcommand{\broken}{\domain}

% Cells
\newcommand{\cell}[1][]{C_{#1}}
\newcommand{\refCell}[1][]{\hat{C}_{#1}}
\newcommand{\cellb}{\partial{} \cell}
\newcommand{\refCellb}{\partial{} \refCell}

% Basis functions
\newcommand{\lagrange}[1][i]{\varphi_{#1}}
\newcommand{\lagrangeRef}[1][i]{\hat{\varphi}_{#1}}
\NewDocumentCommand{\sbasis}{ O{} O{} }{\phi^{#1}_{#2}}
\NewDocumentCommand{\sbasisRef}{ O{} O{} }{\hat{\phi}^{#1}_{#2}} %{\hat{\sbasis[#1][#2]}}
\NewDocumentCommand{\stbasis}{ O{} O{} }{ \psi^{#1}_{#2}}
\NewDocumentCommand{\stbasisRef}{ O{} O{} }{\hat{\psi}^{#1}_{#2}} %{\hat{\stbasis[#1][#2]}}
%\newcommand{\stestfunction}[1]{\sbasis[#1]}
\NewDocumentCommand{\stestfunction}{ O{} O{} }{\sbasis[#1][#2]}
\NewDocumentCommand{\sttestfunction}{ O{} O{} }{\stbasis[#1][#2]}

@@

% \NewDocumentCommand{\stsol}{ O{} O{} }{\overline{u}^{#1}_{#2}}
% \NewDocumentCommand{\stsol}{ O{} O{} }{\underline{\overline{u}}^{#1}_{#2}}

@@

% Integrals
\newcommand{\curTime}{t^k}
\newcommand{\nextTime}{t^{k+1}}
\newcommand{\intdt}[1]{\int_{\curTime}^{\nextTime} #1 \dd{t}}
\newcommand{\intdcell}[1]{\int_{\cell} #1 \dd{\bm{x}}}
\newcommand{\intdrefcell}[1]{\int_{\refCell} #1 \dd{\hat{\bm{x}}}}
\newcommand{\intdcellb}[1]{\int_{\cellb} #1 \dd{S}} % TODO: define boundary of cell
\newcommand{\intdrefcellb}[1]{\int_{\refCellb} #1 \hat{\dd{S}}} % TODO: define boundary of cell

% Yet to be sorted
\newcommand{\quadWeight}[1][i]{w_{#1}}
\newcommand{\quadNode}[1][i]{\hat{x}_{#1}}
\newcommand{\quadNodeNd}[1][i]{\hat{\bm{x}}_{#1}}
\newcommand{\sbasisSize}{(N+1)^\dimensions}
\newcommand{\stbasisSize}{(N+1)^{\dimensions+1}}
\newcommand{\normal}{\bm{n}}

\newcommand{\pleft}{\bm{L}}
\newcommand{\prightsol}{\bm{r}}
\newcommand{\prightpred}{\bm{w}}

\chapter{An ADER-DG scheme for the Navier-Stokes Equations}\label{chap:methods}
This chapter describes the numerical and physical background that is needed to simulate reacting fluids with good accuracy.
We start with a description of the arbitrary high order derivatives (\ader{}) discontinuous Galerkin (\dg{}) method with a focus on equations that contain both advective and diffusive terms.
We then seque into a short description of the \muscl{} scheme which is a second order finite volume scheme.

We conclude this chapter by a description of finite volume limiting and adaptive mesh refinement (\amr{}).
The limiting combines both numerical methods to achieve a stable, high-order method and the \amr{} allows us to perform high quality simulations with a smaller computational cost.

\section{Reactive Compressible Navier Stokes Equations}\label{sec:navier-stokes}
\newcommand{\diffCoeff}{\varepsilon}
\newcommand{\hyperFluxDef}{
  \begin{pmatrix}
    \Qj \\
    \Qv  \otimes \Qj + \bm{I} \pressure  \\
    \Qv \cdot (\bm{I} \QE + \bm{I} \pressure) \\
    \Qj \QZZ
  \end{pmatrix}
}
\newcommand{\viscFluxDef}{
  \begin{pmatrix}
     -\diffCoeff \gradient{\Qrho}\\
     \stressT (\Q, \gradQ)  \\
     \Qv \cdot \stressT (\Q, \gradQ) - \kappa \gradient{T}\\
     -\diffCoeff \gradient{\QZ}
   \end{pmatrix}
}
Fluid motion can be described by the compressible Navier Stokes equations.
We follow the description in~\cite{dumbser2010arbitrary} for the Navier Stokes part.
The coupling with the advection-diffusion-reaction equation follows~\cite{hidalgo2011ader} which describes this equation set for the one-dimensional case.
In the following, we will denote the spatial coordinates as $\bm{x} = \left( x,z \right)$ and $\bm{x} = \left( x,y,z \right)$ for the two and three-dimensional case respectively.
The variable $t$ corresponds to the current physical time.
We solve the \pde{} in the conservative form% \pcref{eq:conservation-law-gradient}
\begin{equation}
 \label{eq:equation-set} 
 \begin{array}{l}
 \text{mass cons.} \\
 \text{momentum cons.} \\
 \text{energy cons.} \\
 \text{cont.\ gas} 
\end{array}
:
\quad
  \pdv{}{t}
  \underbrace{
  \begin{pmatrix}
    \Qrho\\
    \Qj\\
    \QE\\
    \QZ
    \end{pmatrix}}_{\Q}
  +
  \divergence{
  \underbrace{
  \left(
   \underbrace{\hyperFluxDef}_{\hyperFlux(\Q)}
+
\underbrace{\viscFluxDef}_{\viscFlux(\Q, \gradQ)}
  \right)}_{\flux(\Q, \gradQ)}}
 =
  \underbrace{
  \begin{pmatrix}
    \source[\Qrho\phantom{\Qrho}]\\
    \source[\Qj]\\
    \source[\QE]\\
    \source[\QZ]
    \end{pmatrix}}_{\source(\Q, \bm{x}, t)}
\end{equation}
which describes the time evolution of variables $\Q$ with respect to a flux $\flux(\Q, \gradQ)$ and a source $\source(\Q)$.
The vector of conserved quantities is given by
\begin{equation}
  \label{eq:conserved-variables}
 \Q = \left( \Qrho, \Qj, \QE, \QZ \right),
\end{equation}
where $\Qrho$ is the density, $\Qj$ is the two or three-dimensional momentum, $\QE$ the energy density and $\QZ$ is the mass fraction of the chemical reactant.
The chemical reactant in our case is unburnt gas.

The rows of the equation are the conservation of mass, the conservation of momentum, the conservation of energy and the continuity equation of the reactant.
We split the flux into a hyperbolic $\hyperFlux(\Q)$ and a viscous part $\viscFlux(\Q, \gradQ)$
\begin{equation}
  \label{eq:flux}
  \flux(\Q, \gradQ) = \hyperFlux(\Q) + \viscFlux(\Q, \gradQ).
\end{equation}
The hyperbolic flux is given by
\begin{equation}
  \label{eq:hyper-flux}
  \hyperFlux(\Q) = \hyperFluxDef,
\end{equation}
which are the Euler equations coupled with an advection equation.
In this equation $\bm{a} \otimes \bm{b} = \bm{a} \bm{b}^\intercal$ is the Kronecker or outer product of two vectors $\bm{a}$ and $\bm{b}$.
The pressure $\pressure$ is given by the caloric equation of state of an ideal reacting gas
\begin{equation}
  \label{eq:eos}
  \pressure = (\gamma - 1) \left(\QE - \frac{1}{2} \left(\Qv \cdot \Qj \right)  - q_0 \QZ - gz \right).
\end{equation}
The term $q_0 \QZ$ is the chemical energy with the heat release $q_0$, the term $gz$ is the geopotential height.
The pressure is related to the temperature $T$ by the thermic ideal gas law
\begin{equation}
  \label{eq:temperature}
 \frac{\pressure}{\Qrho} = RT,
\end{equation}
where $R$ is the specific gas constant of a fluid.

@@

We also support two different kind of optional source terms.
The first source term is gravity, given by
\begin{equation}\label{eq:source-gravity}
  \source[\Qj] = - g \bm{k} \cdot \Qj,
\end{equation}
where $g = 9.81$ and $\bm{k}$ is a unit vector pointing in $z$-direction\todo{Discuss $g$ either in pressure or scenario.}.

\newcommand{\backgroundPressure}{\overline{\pressure}}
\newcommand{\backgroundRho}{\overline{\Qrho}}
Some of our scenarios can be described as a pertubation of a background state that is initially in hydrostatic balance
\begin{equation}
  \label{eq:hydrostatic-balance}
  \pdv{}{z} \backgroundPressure{\left (z \right )} = -g \backgroundRho(z),
\end{equation}
where $\backgroundPressure$ and $\backgroundRho$ is the background pressure and density.
We now focus on the following part of the momentum equation (neglecting viscous effects) of \cref{eq:equation-set}
\begin{equation}
  \label{eq:momentum-equation}
  \pdv{\Qj}{t}+ \divergence{
    \Qv \cdot \left( \bm{I} \QE + \bm{I} \pressure \right)
  }
  =
  -g \bm{k} \cdot \Qrho.
\end{equation}
The background state \cref{eq:hydrostatic-balance} can be comparitively large and can thus diminish other parts of the flux due to numerical instabilities~\cite{muller2010adaptive}.
To avoid this, we split the pressure as $\pressure = \backgroundPressure(z) + \pressure' (\bm{x}, t)$ and the density as $\Qrho = \backgroundRho(z) + \Qrho'(\bm{x}, t)$.
Inserting this splitting and the definition of divergence into \cref{eq:momentum-equation} allows us to apply \cref{eq:hydrostatic-balance}.
Furthermore, the background states have zero derivatives in all directions but in $z$.
We arrive at 
\begin{equation}
  \label{eq:momentum-equation-split}
  \pdv{\Qj}{t}+ \divergence \left(
    \Qv \cdot \left( \bm{I} \QE + \bm{I} \pressure' \right)
   \right)
  =
  -g \bm{k} \Qrho'.
\end{equation}.
We thus cancel parts of the flux with parts of the source term.
Note that the time derivative of the background density also vanishes.
This form of the equation is inspired by equation set 3 of~\cite{giraldo2008study}.
They additionally use a splitting of the energy and use the pertubations directly as conservative variables.
%We do not use this to simplify the implementation of gradients.

\newcommand{\reactionTimescale}{\tau}
\newcommand{\reactionTemperature}{T_{\text{ign}}}
The chemical reactive source term is
\begin{equation}\label{eq:source-reaction}
  \source[\QZ] = - K(T) \QZ.
\end{equation}
We use the simple reaction model
\begin{equation}\label{eq:reaction-model}
\begin{cases}
  - \frac{1}{\reactionTimescale} & T > \reactionTemperature\\
  0 & \text{otherwise}
\end{cases}  
\end{equation}
where $\reactionTimescale$ is the timescale of the reaction and $\reactionTemperature$ is the activation temperature~\cite{hidalgo2011ader,helzel2000modified}.

\subsection{Boundary conditions}
To close the system we need to impose boundary conditions.\todo{Adapt this chapter to implementation!}

@@

We manually set the energy component of the numerical flux to zero for both no- and free-slip conditions.
\section{The ADER-DG Method}\label{sec:ader-dg}
\newcommand{\NVar}{N^{\text{var}}}
\newcommand{\dimensions}{d}
\newcommand{\mapping}{\mathcal{M}}
\newcommand{\volume}{V}
\newcommand{\cellCenter}{\operatorname{cell-center}}
We describe the arbitrary derivative discontinous Galerkin (\aderdg) method in this chapter.
Our description of the main method follows~\cite{dumbser2008unified,dumbser2010arbitrary,dumbser2018efficient}, our notation follows primarily~\cite{dumbser2018efficient}.
Let $\NVar$ be the number of variables, $\dimensions$ the number of spatial dimensions and $N > 0$ the order of the method.
We discuss the approximation of systems of partial differential equations (\pde) of the form
\begin{equation}
  \label{eq:conservation-law-gradient}
 \frac{\partial}{\partial_t}  \Q + \div \flux(\Q, \gradQ) = \source(\bm{x}, t, \Q),
\end{equation}
which in contrast to hyperbolic conservation laws of the form
\begin{equation}
  \label{eq:conservation-law}
 \frac{\partial}{\partial_t}  \Q + \div \flux(\Q) = \source(\bm{x}, t, \Q)
\end{equation}
contain the gradient $\gradQ \in \mathbb{R}^{\NVar \times \dimensions}$ of the so called vector of conservative variables $\Q \in \mathbb{R}^{\NVar}$~\cite{dumbser2010arbitrary}.
They are therefore not hyperbolic but rather parabolic or elliptic.

We discuss the solution of a conservation law \cref{eq:conservation-law-gradient} with two or three dimensional domain $\domain$ and boundary $\boundary$.
In our implementation of the discontinous Galerkin (\dg) framework, we approximate this solution in the space
\begin{equation}
  \label{eq:dg-space}
  \broken = \bigcup_k \cell[k]
\end{equation}
of disjoint quadrilateral cells $\cell$.
Note that we do not distinguish between the broken approximation space and the domain, the use should be clear given its surrounding context.
For each cell we denote its center by $\cellCenter{\cell}$ and its size by $\Delta x, \Delta y$ and $\Delta z$.

@@

\sidetitle{Reference Cell}
For reasons of computational efficiency, we describe the scheme in terms of a reference cell.
We use the quad cell $\hat{\cell} = (0, 1)^\dimensions$ as space reference cell.
The mapping, which takes a point on the reference cell with coordinates $\hat{\bm{x}}$ and and returns a the coordinates $\bm{x}$ in a grid cell $\cell$ is
\begin{equation}\label{eq:space-mapping}
  \bm{x} = \mapping (\hat{\bm{x}}) =
  \cellCenter(\cell) +
\begin{pmatrix}
\Delta x && \\
&\Delta y & \\
&&\Delta z&
\end{pmatrix}
  \begin{pmatrix}
    \hat{x} - 0.5\\
    \hat{y} - 0.5\\
    \hat{z} - 0.5
  \end{pmatrix}.
\end{equation}
The inverse mapping $\mapping^{-1}$ can be used to define a function $f(\bm{x})$ acting on a point $\hat{\bm{x}}$ of a regular cell in terms of a function $\hat{f}(\bm{x})$ that acts on the corresponding point of the reference cell by the relation
\begin{equation}
  \label{eq:function-ref-cell}
  f(\bm{x}) = f\left( \mapping^{-1}(\bm{x}) \right) = \hat{f}(\hat{\bm{x}}).
\end{equation}

Similar to this, we define a mapping that maps a reference time point $\hat t$ to simulation time $t$
\newcommand{\timeMapping}{\mathcal{T}}
\begin{equation}
  \label{eq:time-mapping}
  t = \timeMapping(\hat{t}) = \curTime + (\Delta t) \hat{t},
\end{equation}
where the reference time interval is $[0,1]$ and the time interval of a cell is $[\curTime, \nextTime]$.
The time at the beginning of a timestep is thus denoted by $\curTime$, the time at the end of a step is denoted by $\nextTime = \curTime + \Delta t$ with timestep size $\Delta t$.

The Jacobian determinant of $\mapping$ is simply the volume $V$ of the cell
\begin{equation}
  \label{eq:determinant-mapping}
  \volume = \operatorname{det}(\operatorname{Jacobian}\mapping) = \Delta x \, \Delta y \,\Delta z,
\end{equation}
the Jacobian determinant of $\timeMapping$ is $\Delta t$.
This is useful for evaluating derivatives by the chain rule and, similarly, integrals by substitution.
For example, we can evaluate a volume integral over a cell by
\begin{equation}
  \label{eq:integration-by-substitution}
  \intdcell{
f(\bm{x})
  }
  =
\volume \intdrefcell{
    \hat{f}(\hat{\bm{x}})
  }
\end{equation}
\sidetitle{Basis Functions}
The next step is the definition of basis functions for our cells.
We use Langrange polynomials that are defined in the one-dimensional case for the reference interval $(0,1)$ by
\begin{align}
  \label{eq:lagrange-basis}
    \lagrangeRef(x) = \prod_{i >0, j>0, i \neq n}^N \frac{x - {\quadNode[j]}}{x_i - \quadNode[j]},
\end{align}
We can then approximate functions as a sum over the support points $\quadNode[j]$
\begin{equation}
  \label{eq:langrange-expansion-1d}
  f(x) = \sum_{i = 0}^N f(x_i) \lagrangeRef[i](x) = f(x_i) \lagrangeRef[i](x),
\end{equation}
where we used the Einstein summation convention which implies summation over repeated free indices.
This convention is used from here on.
The support points $\quadNode[i]$ are chosen such that they coincide with the nodes of the Gauss-Legendre quadrature of order $N+1$ which we can use to evaluate integrals
\begin{equation}
  \label{eq:quadrature-1d}
  \int_0^1 f(\hat{x}) \dd{\hat{x}} = \sum_{i=0}^{N} \quadWeight[i] f(\quadNode[i]),
\end{equation}
where $\quadWeight[i]$ is the quadrature weight corresponding to the node $\quadNode[i]$.

The basis functions are similarly defined in $\dimensions$-dimensions by considering nodes that are the tensor-products of $d$ one-dimensional Gauss-Legendre quadrature.
They are indexed by a multi-index $n = (n_1, \ldots, n_\dimensions)$
\begin{equation}
  \label{eq:basis-space}
  \sbasisRef[][n] (\bm{x}) = \prod_{i=0}^{\dimensions} \lagrangeRef[n_i] (x_i)
\end{equation}
that is linearised such that $n = (0, \ldots, \sbasisSize)$.
A $d$-dimensional function is then approximated by
\begin{equation}
  \label{eq:lagrange-expansion}
  f(\bm{x}) = \sum_{i = 0}^{\sbasisSize - 1} f(\bm{x}_i) \lagrangeRef[i](\bm{x}) = f(\bm{x}_i) \lagrangeRef[i](\bm{x}).
\end{equation}
Similarly to the one-dimensional case, we can evaluate integrals by
\begin{align}
  \label{eq:quadrature}
  \underbrace{\int_0^1 \cdots \int_0^1}_{\dimensions \text{ times}} f(\hat{\bm{x}}) \dd{\hat{\bm{x}}} &=
 \sum_{n=0}^{\sbasisSize-1} \quadWeight[n] f(\quadNodeNd[n]) \\
  \intertext{with}
  \quadWeight[n] &= \prod_{i}^{d} \quadWeight[n_i],
\end{align}
where the quadrature weights $\quadWeight[n]$ are the product of all relevant one-dimensional quadrature weights, using the aforementioned (linearized) multi-index $n$.
\todo{Should be correct. Is there a way to avoid this index hell, esp.\ multi-index?}

### 2019-01-11 writing/thesis/pages/abstract.tex
Change stuff.
Update methods.

We use the compressible Navier-Stokes (NS) equations with heat conduction to describe the dynamics of clouds.
Due to the diffusive components, the NS-Equations are not hyperbolic.
We thus need to modify parts of the numerical scheme.

>>>>

We use the compressible Navier-Stokes (NS) equations with heat conduction to simulate the dynamics of clouds.
Due to the diffusive components, the NS-Equations are not hyperbolic.
We thus need to modify parts of the numerical scheme.

Furthermore, we extend the equation set to the reactive compressible Navier-Stokes equations which include a source term that models chemical reactions.
